%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[a4paper,oneside,12pt]{article}
\usepackage{mystyle}

\begin{document}

\title{\Large\bf Integers}
\author{%%
  Minh Van Nguyen \\
  \url{mvngu@gmx.com}
}
\date{\today}
\maketitle


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Even and odd}

A \emph{set} is a bunch or collection of objects.  The objects of a
set are called its \emph{elements}, which must all be different.  For
example, the set of all rubbish bins at your house, the set of all
fingers on your two hands, and the set of all whole numbers from $1$
to $10$.  You use curly braces to represent a set.  The set of all
even whole numbers from $1$ to $10$ can be written as
$\set{\quintuple{2}{4}{6}{8}{10}}$, the set of whole numbers from $1$
to $5$ can be written as $\set{\quintuple{1}{2}{3}{4}{5}}$, and the
set of the first four prime numbers is $\set{\quadruple{2}{3}{5}{7}}$.
The set $\set{\triple{1}{2}{1}}$ is the same as the set
$\set{\pair{1}{2}}$ because the elements of a set must all be
distinct.

\begin{exercise}
Write out the set of all whole numbers from $1$ to $5$.
\end{exercise}

\ifbool{showSolution}{
\begin{solution}
The set of all whole numbers from $1$ up to and including $5$ can
be written as $\set{\quintuple{1}{2}{3}{4}{5}}$.
\end{solution}
}{}

\begin{exercise}
Give two more examples of sets.  The sets do not have to be
collections of numbers.
\end{exercise}

\ifbool{showSolution}{
\begin{solution}
The set of all pens in your pencil case.  The set of all cars that you
own.  If you do not own a car, then this set of cars has zero
elements.  The set that has zero elements is called the \emph{empty}
set and is written as $\emptyset$.
\end{solution}
}{}

An \emph{integer} is a whole number such as $3$, $7$, $\the\year$,
$-5$, or $-9$.  The set of all integers is written as
$\ZZ = \set{\dots,\, -3,\, -2,\, -1,\, 0,\, 1,\, 2,\, 3,\dots}$.  You
write $2 \in \ZZ$ to mean that the number $2$ belongs to the set of
integers.

\begin{definition}
\textbf{Even integers.}
An integer $n$ is \emph{even} if you can write it as $n = 2k$, where
$k$ is some other integer.
\end{definition}

\begin{definition}
\textbf{Odd integers.}
An integer $n$ is \emph{odd} if you can write it as $n = 2r + 1$,
where $r$ is some other integer.
\end{definition}

For example, the integer $2$ is even so you can write
$2 = 2 \times 1$.  The integer $6$ is even so you can write
$6 = 2 \times 3$.  The integer $3$ is odd and you can write it as
$3 = 2 \times 1 + 1$.  The integer $9$ is odd because
$9 = 2 \times 4 + 1$.

\begin{exercise}
Is the integer $32$ even?  If yes, write $32$ in the form $2k$, where
$k$ is an integer.  If no, explain why $32$ is not even.
\end{exercise}

\ifbool{showSolution}{
\begin{solution}
The integer $32$ is even because it can be written as
$32 = 2 \times 16$.
\end{solution}
}{}

\begin{exercise}
Is the integer $51$ even?  If yes, explain why.  If no, explain why
not.
\end{exercise}

\ifbool{showSolution}{
\begin{solution}
The integer $51$ is odd because you can write $51$ as
$51 = 2 \times 25 + 1$.
\end{solution}
}{}

The following rules are very useful when you want to simplify a
complicated mathematical expression.

\begin{definition}
\textbf{Associative laws.}
If $\triple{a}{b}{c}$ are any numbers, then the
\emph{associative laws} state that you can write:
%%
\begin{packedenumeral}
\item $abc = a (bc) = (ab) c$

\item $a + b + c = a + (b + c) = (a + b) + c$
\end{packedenumeral}
\end{definition}

This means that for the product $abc$, you can first simplify $bc$.
If you want, you can simplify $ab$ first.  Similarly, for the sum
$a + b + c$, you can simplify $a + b$ first.  If you want, you can
simplify $b + c$ first.  Now let's put the associative laws to some
good use.

\begin{exercise}
Simplify the product $2 \times 3 \times 5$ in two ways.
\end{exercise}

\ifbool{showSolution}{
\begin{solution}
Using the associative laws, you can simplify $3 \times 5$ first so
you can write
%%
\begin{align*}
2 \times 3 \times 5
&=
2 \times (3 \times 5) \\[4pt]
&=
2 \times 15 \\[4pt]
&=
30.
\end{align*}
%%
On the other hand, the associative laws also allow you to simplify
$2 \times 3$ first.  In other words, you can also write
%%
\begin{align*}
2 \times 3 \times 5
&=
(2 \times 3) \times 5 \\[4pt]
&=
6 \times 5 \\[4pt]
&=
30.
\end{align*}
\end{solution}
}{}

\begin{exercise}
Simplify the sum $3 + 5 + 7$ in two ways.
\end{exercise}

\ifbool{showSolution}{
\begin{solution}
The associative laws allow you to first simplify $3 + 5$ and so you
can write
%%
\begin{align*}
3 + 5 + 7
&=
(3 + 5) + 7 \\[4pt]
&=
8 + 7 \\[4pt]
&=
15.
\end{align*}
%%
Furthermore, the associative laws also allow you to first simplify
$5 + 7$.  This means that you can also write
%%
\begin{align*}
3 + 5 + 7
&=
3 + (5 + 7) \\[4pt]
&=
3 + 12 \\[4pt]
&=
15.
\end{align*}
\end{solution}
}{}

\begin{theorem}
\label{thm:product_of_two_evens_is_even}
If $n$ and $r$ are even integers, then $nr = n \times r$ is even.
\end{theorem}

\begin{proof}
Since $n$ and $r$ are even, then you can write each of them as
$n = 2k$ and $r = 2s$, where $k$ and $s$ are integers.  Now you can
use the associative laws to write the product $nr$ as
%%
\begin{align*}
nr
&=
(2k) \times (2s) \\[4pt]
&=
2 \times k \times 2 \times s \\[4pt]
&=
2 \times (k \times 2 \times s) \\[4pt]
&=
2 (2ks).
\end{align*}
%%
Whatever the number $2ks$ is, it will be an integer because the
multiplication of any three integers is an integer.  Make the
substitution $t = 2ks$ and you have $nr = 2t$, where $t = 2ks$ is an
integer.  Therefore $nr$ is even.
\end{proof}

You can use \Theorem{thm:product_of_two_evens_is_even} to conclude
whether the multiplication of two integers is even.  For example, you
don't have to simplify the product $6 \times 8$ to know that the
result will be an even integer.  Since $6 = 2 \times 3$ and
$8 = 2 \times 4$, then you can write
%%
\begin{align*}
6 \times 8
&= (2 \times 3) \times (2 \times 4) \\[4pt]
&=
2 \times (3 \times 2 \times 4)
\end{align*}
%%
which is an even integer.  \Theorem{thm:product_of_two_evens_is_even}
also states that the multiplication of $2$ by any integer will be
even.  If you multiply $2$ by a million, you don't have to work out
the result to know that the result will be an even number.

The following rules can also help you to simplify a complicated
mathematical expression.

\begin{definition}
\textbf{Distributive laws.}
If $\quadruple{a}{b}{c}{d}$ are any numbers, the
\emph{distributive laws} state that you can write:
%%
\begin{packedenumeral}
\item $a (b + c) = ab + ac$

\item $(a + b) \times (c + d) = a(c + d) + b(c + d)$
\end{packedenumeral}
\end{definition}

The distributive laws say that if you have any three numbers
$\triple{a}{b}{c}$ then you can simplify the expression
$a(b + c)$ in two ways.  First, you can simplify $b + c$ and then
multiply the result by $a$.  Or you can simplify $ab$ and $ac$ and
then add the two together.  Suppose you want to simplify the
expression $3 (2 + 7)$.  You can first simplify the addition and then
multiply the result by $3$:
%%
\begin{align*}
3 (2 + 7)
&=
3 \times 9 \\[4pt]
&=
27.
\end{align*}
%%
On the other hand, the distributive laws also allow you to write the
expression $3 (2 + 7)$ as $3 (2 + 7) = (3 \times 2) + (3 \times 7)$.
Now you simplify the multiplications and then add the result together:
%%
\begin{align*}
3 (2 + 7)
&=
(3 \times 2) + (3 \times 7) \\[4pt]
&=
6 + 21 \\[4pt]
&=
27.
\end{align*}

\begin{exercise}
How would you simplify the expression $5 (3 + 1)$ in two ways?
\end{exercise}

\ifbool{showSolution}{
\begin{solution}
First, you can simplify the expression $5 (3 + 1)$ by simplifying the
sum $3 + 1$ and then multiply the result by $5$.  This will result in
%%
\begin{align*}
5 (3 + 1)
&=
5 \times 4 \\[4pt]
&=
20.
\end{align*}
%%
Second, you can use the distributive laws to write
%%
\begin{align*}
5 (3 + 1)
&=
(5 \times 3) + (5 \times 1) \\[4pt]
&=
15 + 5 \\[4pt]
&=
20.
\end{align*}
\end{solution}
}{}

The distributive laws also say that if you have any four numbers
$\quadruple{a}{b}{c}{d}$ then you can simplify the expression
$(a + b) \times (c + d)$ in two ways.  First, you can simplify each of
$a + b$ and $c + d$ and then multiply the two results together.
Second, you can simplify each of $a(c + d)$ and $b(c + d)$ and then
add the two results.  For example, you can simplify the expression
$(2 + 3) \times (5 + 7)$ in two ways as follows.  The first way is:
%%
\begin{align*}
(2 + 3) \times (5 + 7)
&=
5 \times 12 \\[4pt]
&=
60.
\end{align*}
%%
The second way is:
%%
\begin{align*}
(2 + 3) \times (5 + 7)
&=
2(5 + 7) + 3(5 + 7) \\[4pt]
&=
(2 \times 12) + (3 \times 12) \\[4pt]
&=
24 + 36 \\[4pt]
&=
60.
\end{align*}

\begin{exercise}
How would you simplify the expression $(1 + 2) \times (3 + 4)$ in two
ways?
\end{exercise}

\ifbool{showSolution}{
\begin{solution}
First, you can simplify the sums inside the parentheses and then
multiply the results together.  This will result in
%%
\begin{align*}
(1 + 2) \times (3 + 4)
&=
3 \times 7 \\[4pt]
&=
21.
\end{align*}
%%
On the other hand, you can use the distributive laws to simplify the
expression as follows:
%%
\begin{align*}
(1 + 2) \times (3 + 4)
&=
1(3 + 4) + 2(3 + 4) \\[4pt]
&=
(1 \times 7) + (2 \times 7) \\[4pt]
&=
7 + 14 \\[4pt]
&=
21.
\end{align*}
\end{solution}
}{}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Rational and irrational numbers}

A \emph{rational} number is a ratio of two integers.  A
rational number is also known as a fraction.  These are rational
numbers: $1/2$, $3/4$, $11/9$, and $-10/2$.  You write $5/2 \in \QQ$
to mean that $5/2$ belongs to the set of rational numbers.  The set of
all rational numbers is written as
\[
\QQ
=
\setdes{
  \frac{a}{b}
}{
  \pair{a}{b} \in \ZZ
  \text{ and }
  b \neq 0
}.
\]
This means that the set of rational numbers consists of all numbers of
the form $a/b$ such that $a$ and $b$ are integers, but $b$ cannot be
zero.  The top integer $a$ is called the \emph{numerator} and the
bottom integer $b$ is called the \emph{denominator}.  If $b = 1$, then
$a/b = a/1 = a$ and consequently any integer is also a rational
number.

\begin{exercise}
Provide an example of a rational number that is also an integer.
Explain why your number is an integer.
\end{exercise}

\ifbool{showSolution}{
\begin{solution}
The rational number $2/1$ is also an integer because it can be written
as $2/1 = 2$ and $2$ is an integer.
\end{solution}
}{}

\begin{exercise}
Give an example of a rational number that is not an integer.  Explain
why your number is not an integer.
\end{exercise}

\ifbool{showSolution}{
\begin{solution}
The ratio $1/2$ is not an integer because $1/2$ cannot be simplified
to any integer.
\end{solution}
}{}

\begin{exercise}
If $a/b$ is a rational number, why can't $b$ be zero?  Consider the
examples of $1/4$ and $1/0$.
\end{exercise}

\ifbool{showSolution}{
\begin{solution}
If in the rational number $a/b$ you have $b = 0$, then the ratio
$a / 0$ is not defined.  This means that as $b$ comes closer and
closer to zero, the value of the ratio $a/b$ does not come to a fixed
number.
\end{solution}
}{}

An \emph{irrational} number is any number that cannot be written as a
ratio of two integers.  The number $\sqrt{2}$ is an irrational number
so it cannot be an integer and therefore the expression
$\displaystyle{\frac{\sqrt{2}}{2}}$ is not a fraction.

\begin{exercise}
Provide another example of an irrational number.
\end{exercise}

\ifbool{showSolution}{
\begin{solution}
The number $\pi = 3.141592\dots$ is an irrational number.
\end{solution}
}{}

The set of \emph{real} numbers is made up of all rational and all
irrational numbers.  The set of real numbers is written as $\RR$.  For
example, the ratio $3/7$ is a real number, the square root $\sqrt{2}$
is a real number, and the integer $42$ is a real number.

\begin{exercise}
Explain why any integer is a real number.
\end{exercise}

\ifbool{showSolution}{
\begin{solution}
Any integer is also a rational number.  Since a rational number is
also a real number, it follows that any integer is also a real
number.
\end{solution}
}{}

\begin{exercise}
How can a number be represented as a picture?
\end{exercise}

\ifbool{showSolution}{
\begin{solution}
A real number can be represented as a point on the number line.
\end{solution}
}{}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Order of operations}

There are four main operations that are defined on real numbers.
These operations are summarised below:
%%
\begin{packedenumeral}
\item Addition is when you add two numbers together.  If $a$ and $b$
  are real numbers, you write $a + b$ to mean that you want to sum $a$
  and $b$.

\item Subtraction is finding the difference between one number and
  another number.  You write $a - b$ to mean that you subtract $b$
  from $a$.

\item Multiplication is when you multiply two numbers together.  The
  product of $a$ and $b$ can be written as $a \times b$ or $ab$.

\item Division is finding the ratio of two numbers.  The ratio of $a$
  over $b$ is written as the fraction $\frac{a}{b}$ or $a/b$, but $b$
  must not be zero.
\end{packedenumeral}

When you want to simplify a complicated expression, you should follow
a number of rules to help you.  These rules are called the
\emph{order of operations} and are listed below in the order in which
they should be followed.
%%
\begin{packedenumeral}
\item If numbers are grouped together within parentheses, you start
  from the innermost parentheses and work outward.

\item If there are multiplications and/or divisions, work from left to
  right to simplify the multiplications and/or division.

\item If there are additions and/or subtractions, work from left to
  right to simplify the additions and/or subtractions.
\end{packedenumeral}

\begin{exercise}
Let $p$ denote Patrick's age, $m$ be Marcus' age, and $u$ be your
uncle's age.  The sum of all these ages is $p + m + u$.  How would you
simplify this expression?
\end{exercise}

\ifbool{showSolution}{
\begin{solution}
Working from left to right, you can write $(p + m) + u$ to indicate
that you want to first simplify the sum $p + m$.  Use your result to
then simplify $(p + m) + u$.  On the other hand, you can write
$p + (m + u)$ to mean that you want to first simplify the sum
$m + u$.  You then use this result to simplify the expression
$p + (m + u)$.
\end{solution}
}{}

\begin{exercise}
Simplify the expression $3 + 2 \times 4 - 8/4 + 9$.
\end{exercise}

\ifbool{showSolution}{
\begin{solution}
The expression $3 + 2 \times 4 - 8/4 + 9$ can be simplified as
%%
\begin{align*}
3 + 2 \times 4 - 8/4 + 9
&=
3 + 8 - 8/4 + 9 \\[4pt]
&=
3 + 8 - 2 + 9 \\[4pt]
&=
11 - 2 + 9 \\[4pt]
&=
9 + 9 \\[4pt]
&=
18.
\end{align*}
\end{solution}
}{}

\begin{exercise}
Simplify the expression
$\displaystyle{
  \frac{
    3 \times 2 + (11 - 5)
  }{
    2 \times (3 + 7)
  }
}$.
\end{exercise}

\ifbool{showSolution}{
\begin{solution}
The expression
$\displaystyle{
  \frac{
    3 \times 2 + (11 - 5)
  }{
    2 \times (3 + 7)
  }
}$
can be simplified as
%%
\begin{align*}
\frac{
  3 \times 2 + (11 - 5)
}{
  2 \times (3 + 7)
}
&=
\frac{
  3 \times 2 + 6
}{
  2 \times (3 + 7)
} \\[4pt]
&=
\frac{
  3 \times 2 + 6
}{
  2 \times 10
} \\[4pt]
&=
\frac{
  6 + 6
}{
  20
} \\[4pt]
&=
\frac{
  12
}{
  20
} \\[4pt]
&=
\frac{
  3
}{
  5
}.
\end{align*}
\end{solution}
}{}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Problems}

\begin{problem}
\item What sort of things can be represented as positive integers?
\ifbool{showSolution}{
  \begin{solution}
  Money, the number of hair on your head~(if you are not bald), the
  number of toes on your feet~(if you have toes), the number of houses
  in Melbourne.
  \end{solution}
}{}

\item What sort of things can be represented as negative integers?
\ifbool{showSolution}{
  \begin{solution}
  The temperature below zero can be represented as a negative
  integer.
  \end{solution}
}{}

\item Explain why the integer $-42$ is even.
\ifbool{showSolution}{
  \begin{solution}
  The integer $-42$ is even because the number can be written as
  $-42 = 2 \times (-21)$.
  \end{solution}
}{}

\item Give an example of something that can be represented as a
  rational number.
\ifbool{showSolution}{
  \begin{solution}
  The number of oranges you have eaten divided by the number of
  oranges you have altogether.
  \end{solution}
}{}

\item Provide an example of something that can be represented as a
  real number.
\ifbool{showSolution}{
  \begin{solution}
  The area of your house.
  \end{solution}
}{}

\item Simplify the expression
  \[
  E
  =
  \frac{
    3 m c^2
  }{
    (6 - 4) + 1
  }.
  \]
\ifbool{showSolution}{
  \begin{solution}
  \begin{align*}
  E
  &=
  \frac{
    3 m c^2
  }{
    (6 - 4) + 1
  } \\[4pt]
  &=
  \frac{
    3 m c^2
  }{
    2 + 1
  } \\[4pt]
  &=
  \frac{
    3 m c^2
  }{
    3
  } \\[4pt]
  &=
  m c^2.
  \end{align*}
  \end{solution}
}{}

\item The number $e = 2.71828\dots$ is called \emph{Euler's constant}.
  Read about $e$ on Wikipedia or search on the Internet for
  ``Euler's constant.''  Is $e$ a rational number?  Is $e$ an
  irrational number?  Is $e$ an integer?  Does $e$ belong to the set
  $\RR$ of real numbers?  Where can you find a use for the number $e$?
\ifbool{showSolution}{
  \begin{solution}
  Euler's constant $e = 2.71828\dots$ is an irrational number, which
  means that $e$ also belongs to the set of real numbers.  The number
  $e$ is used as the base of the \emph{natural logarithm},
  i.e.~logarithm to the base $e$.  The number $e$ is also used in the
  calculation of compound interest.
  \end{solution}
}{}

\item Write out the set of the first five prime numbers.
\ifbool{showSolution}{
  \begin{solution}
  The set of the first five prime numbers is
  $\set{\quintuple{2}{3}{5}{7}{11}}$.
  \end{solution}
}{}

\item Explain what is wrong with the following simplification:
  %%
  \begin{align*}
  2 \times (1 + 3)
  &=
  2 \times 1 + 3 \\[4pt]
  &=
  2 + 3 \\[4pt]
  &=
  5.
  \end{align*}
\ifbool{showSolution}{
  \begin{solution}
  The addition inside the parentheses should be simplified first.  The
  simplification should be:
  %%
  \begin{align*}
  2 \times (1 + 3)
  &=
  2 \times 4 \\[4pt]
  &=
  8.
  \end{align*}
  \end{solution}
}{}

\item Is the square root $\sqrt{4 \times 16}$ an integer, a rational
  number, or an irrational number?
\ifbool{showSolution}{
  \begin{solution}
  The square root $\sqrt{4 \times 16}$ can be simplified as
  %%
  \begin{align*}
  \sqrt{4 \times 16}
  &=
  \sqrt{4} \times \sqrt{16} \\[4pt]
  &=
  2 \times 4 \\[4pt]
  &=
  8.
  \end{align*}
  %%
  Therefore $\sqrt{4 \times 16}$ is an integer.
  \end{solution}
}{}

\item Let $x$ be any even integer and $y$ any odd integer.  Explain
  why the product $xy = x \times y$ is an even integer.
\ifbool{showSolution}{
\begin{solution}
You are given that $x$ is an even integer so you can write $x$ as
$x = 2k$, where $k$ is another integer.  You also know that $y$ is an
odd integer so you can write it as $y = 2r + 1$, where $r$ is another
integer.  You can now use the distributive and associative laws to
write the product $xy$ as
%%
\begin{align*}
xy
&=
(2k) \times (2r + 1) \\[4pt]
&=
(2k \times 2r) + (2k \times 1) \\[4pt]
&=
2 (2kr) + 2k \\[4pt]
&=
2 (2kr + k).
\end{align*}
%%
The number $2 (2kr + k)$ is even because the product of $2$ by any
integer is even.  Therefore $xy$ is even.
\end{solution}
}{}

\item This project requires you to measure the amount of time that you
  spend showering.  For example, suppose you will shower this Saturday
  evening.  Before you turn on the shower water, write down the
  current date and time~(e.g.~Saturday~13th~January~2018, start at
  5:30~pm).  After you have showered and turned off the shower water,
  write down the time that you turned off the water~(e.g.~end at
  5:37~pm).  The difference between these two times is the amount of
  time that you spent showering during that evening.  In the above
  example times, the amount of time spent showering would be seven
  minutes.  Record the amount of time you spent showering for each day
  that you shower.  It's like keeping a shower diary.  Bring along
  your shower diary next week and show me.
\end{problem}

\end{document}
